
# Supplementary Tables and Figures.

\newpage

(ref:tableS1-caption) Skills and skill change in the two samples (after weighting).

```{r tableS1, echo = FALSE}
# Merge information about change in literacy and numeracy per study
# i.e., combine the information for both domains
# Compute pooled correlations between change (delta) and initial level (t1)
cors_pooler2 <- function(target) {
  data_filtered <- get(target) %>%
    split(.$.imp) %>%
    imputationList()

  sink("NUL")

  results <- with(data_filtered, lm(scale(t2_pv - t1_pv) ~ scale(t1_pv),
    weights = weight
  )) %>%
    MIcombine() %>%
    summary()

  sink()

  resultlist <- tibble(
    rho = results[["results"]][2],
    se = results[["se"]][2],
    lower = results[["(lower"]][2],
    upper = results[["upper)"]][2]
  )

  resultlist
}

# Create a tibble for computing the PV-based correlations of t1 with delta
cors_t1_delta <- tibble(
  target = c("reading_neps", "math_neps", "reading_piaac", "math_piaac"),
)

# Map the function to the tibble to obtain the results
cors_t1_delta <- cors_t1_delta %>%
  rowwise() %>%
  mutate(ergebnis = list(cors_pooler2(target))) %>%
  unnest(ergebnis) %>%
  mutate(
    domain = str_extract(target, pattern = "reading|math"),
    study = str_extract(target, pattern = "neps|piaac"),
    CI = str_c("[", round(lower, 2), "; ", round(upper, 2), "]")
  )

# Combine descriptive statistics about change in a reduced and formatted table
skills_descriptives <- deltas %>%
  filter(grouping == "total") %>%
  select(target, t1, sd_t1, t2, sd_t2, delta, sd_delta) %>%
  arrange(-grepl("piaac", target)) %>%
  pivot_longer(!target) %>%
  mutate(
    study = str_extract(target, "piaac|neps"),
    domain = str_extract(target, "math|reading")
  ) %>%
  select(-target) %>%
  pivot_wider(names_from = c(study, domain), values_from = c(value)) %>%
  mutate(name = str_replace_all(
    name,
    c(
      "sd_t[12]|sd_delta" = "\\\\emph{SD}",
      "t[12]_{0}|delta" = "\\\\emph{M}"
    )
  )) %>%
  select(
    Parameter = name,
    "Literacy" = piaac_reading,
    "Numeracy" = piaac_math,
    "Reading" = neps_reading,
    "Math" = neps_math
  )

apa_table(skills_descriptives,
  col_spanners =
    list("\\makecell[c]{PIAAC-L\\\\(3 years)}" = c(2, 3), "\\makecell[c]{NEPS\\\\(6 years)}" = c(4, 5)),
  stub_indents = list(
    "Skill level at $T_1$" = c(1:2),
    "Skill level at $T_2$" = c(3:4),
    "Skill change ($\\Delta{T_1, T_2}$)" = c(5:6)
  ),
  escape = F,
  caption = "(ref:tableS1-caption)",
  font_size = "small",
  note = "Means are in the raw metric of the tests (0-500 points in PIAAC; logits in NEPS). All statistics are pooled across 10 (20) PVs in PIAAC-L (NEPS)."
)
```
\newpage

(ref:tableS2-caption) Linear age differences in literacy and numeracy skills as estimated from the cross-sectional ($T_1$) data.

```{r tableS2, echo = F, warning = F}
# Table is sourced from 04_reporting.R
# source("04_reporting.R")

age_effects %>%
  mutate(Change = str_replace_all(
    Change,
    "SD",
    "\\\\emph{SD}"
  )) %>%
  rename("Linear age effect" = Change) %>%
  apa_table(
    col.names = c("Linear age effect", rep(c("Literacy", "Numeracy"), 2)),
    col_spanners =
      list(
        "PIAAC-L" = c(2, 3),
        "NEPS" = c(4, 5)
      ),
    stub_indents = list(
      "For one year" = c(1:2),
      "For the study period" = c(3:4),
      "For one decade" = c(5:6)
    ),
    escape = F,
    caption = "(ref:tableS2-caption)",
    align = "lcccc",
    font_size = "small",
    note = "Cross-sectional age differences in skills  estimated from a linear model regressing $T_1$ skill level on age. The study period was three years in PIAAC-L and six years in NEPS. \"$SD$\" refers to change expressed in pooled standard deviations of the respective skill. All estimates are pooled across the 10 (20 in NEPS) PVs."
  )
```

\newpage

(ref:figureS1-caption) Cross-sectional age profiles of literacy and numeracy skills at $T_1$. The horizontal line represents the sample mean. The *Y* axis shows the raw metric of the tests (0-500 points in PIAAC; logits in NEPS).

```{r figureS1, echo = FALSE, warning = FALSE, fig.dim = c(8,7), fig.cap = "(ref:figureS1-caption)"}

ageplot <- function(data) {

  # Compute skill mean, min and max at T1 - then pool the data across PVs

  plotdata <- get(data)
  t1skills <- plotdata %>%
    group_by(.imp) %>%
    summarise(
      Mean = mean(t1_pv),
      Min = min(t1_pv),
      Max = max(t1_pv)
    ) %>%
    summarise(across(where(is.numeric), mean))


  domain <- if_else(grepl("reading", data), "Literacy", "Numeracy")

  plotlims <- list(
    upper = ifelse(grepl("piaac", data),
      300, 1
    ),
    lower = ifelse(grepl("piaac", data),
      250,
      -1
    )
  )

  # Plot the age profile based on the first PV
  ageplot <- plotdata %>%
    filter(.imp == 1) %>%
    ggplot(aes(x = age, y = t1_pv)) +
    expand_limits(
      y = c(plotlims$lower, plotlims$upper),
      x = c(18, 67)
    ) +
    geom_smooth(
      # span = 0.1, level = 0.95,
      fill = "#005B96", color = "#005B96"
    ) +

    # stat_smooth(level = 0.99, span = 0.01) +
    geom_hline(aes(yintercept = t1skills[["Mean"]]), color = "#666666") +
    xlab(expression(paste("Age at ", T[1], " in years"))) +
    ylab(expression(paste("Skill level at ", T[1]))) +
    # scale_y_continuous(breaks =
    #                     ifelse(grepl("piaac", data),
    #                     seq(250, 300),
    #                     seq(-1, 1))
    #                     ) +
    theme(
      text = element_text(family = "Source Sans Pro"),
      panel.background = element_rect(fill = "white", color = "black"),
      axis.title.x = element_text(size = 10, margin = margin(t = 10)),
      axis.title.y = element_text(size = 10, margin = margin(r = 10)),
      axis.text = element_text(size = 10),
      axis.line = element_line(color = "black"),
      panel.grid.major = element_line(colour = "grey90", size = 0.1),
      plot.margin = margin(1, 0.5, 0.5, 0.5, "cm")
    )

  ageplot
}

plot_grid(
  ageplot("reading_piaac"),
  ageplot("reading_neps"),
  ageplot("math_piaac"),
  ageplot("math_neps"),
  labels = c(
    "Literacy (PIAAC-L)", "Literacy (NEPS)",
    "Numeracy (PIAAC-L)", "Numeracy (NEPS)"
  ),
  label_size = 12,
  label_fontfamily = "Source Sans Pro",
  label_y = 1
)
```

\newpage

(ref:figureS2-caption) Distribution of the change scores ($\Delta{T_1, T_2)}$. The dashed vertical line represents "no change". The dark blue areas represent change greater than 0.8 *SD*, which is conventionally seen as a "large" effect. 

```{r figureS2, echo = FALSE, warning = FALSE, fig.dim = c(8,7), fig.cap = "(ref:figureS2-caption)"}

delta_density <- function(data) {
  data <- get(data)
  # Compute sd at T1 (for computation of Cohen's dav and for coloring the plot)
  require(radiant.data)

  sds <- data %>%
    group_by(.imp) %>%
    summarise(
      sd_pooled = 0.5 * (weighted.sd(t1_pv, weight) +
        weighted.sd(t2_pv, weight)),
      sd_delta = weighted.sd(t2_pv - t1_pv, weight)
    ) %>%
    summarise(sd_pooled = mean(sd_pooled), sd_delta = mean(sd_delta))

  # Compute deltas (t2-t1) based on the EAP scores
  plotdata <- data %>% ungroup() %>%
    # group_by(seqid) %>%
    mutate(delta = t1_pv - t2_pv)

  # Compute densities for each delta value and categorize them according to  whether they exeed a "large" effect.
  plotdata <- with(density(plotdata$delta), tibble(x, y)) %>%
    mutate(
      direction = ifelse(x <= -0.8 * sds[["sd_pooled"]], "neg",
        ifelse(x >= 0.8 * sds[["sd_pooled"]], "pos", "little")
      )
    )

  # Create a density plot
  plotted <- plotdata %>% ggplot(aes(x = x, y = y)) +
    geom_area(aes(fill = direction),
      alpha = 1,
      show.legend = F
    ) +
    geom_vline(
      xintercept = 0,
      size = 0.5,
      lty = "dashed",
      color = "#666666"
    ) +
    geom_line() +
    theme_apa() +
    xlab(expression(paste("Change score (", Delta, "T2, T1", ")"))) +
    ylab("Density") +
    theme(
      text = element_text(family = "Source Sans Pro"),
      axis.title = element_text(size = 10),
      axis.text = element_text(size = 10),
      plot.margin = margin(1, 0.5, 0.5, 0.5, "cm")
    ) +
    scale_fill_manual(values = c("#BDCDD8", "#005B96", "#005B96")) +
    expand_limits(x = c(-2 * sds$sd_pooled, 2 * sds$sd_pooled))

  return(plotted)
}

plot_grid(
  delta_density("reading_piaac"),
  delta_density("reading_neps"),
  delta_density("math_piaac"),
  delta_density("math_neps"),
  labels = c(
    "Literacy (PIAAC-L)", "Literacy (NEPS)",
    "Numeracy (PIAAC-L)", "Numeracy (NEPS)"
  ),
  label_size = 12,
  label_fontfamily = "Source Sans Pro",
  label_y = 1
)
```

\newpage

(ref:figureS3-caption) Scatterplots of literacy and numeracy skills at the two time points. For better legibility, the plot shows data only from the first five plausible values (PV). Axes are in the original metrics of the PIAAC (0–500 points) and NEPS tests (logits), respectively. The lines show the linear correlation coefficient $r_{T_1, T_2}$ based on each of the five PVs.

```{r figureS3, echo = FALSE, warning = FALSE, fig.dim = c(9,7), fig.cap = "(ref:figureS3-caption)"}

point_plotter <- function(data) {
  y_lab <- if_else(grepl("reading", data),
    expression(paste("Literacy skills at  ", T[2])),
    expression(paste("Numeracy skills at  ", T[2]))
  )
  x_lab <- if_else(grepl("reading", data),
    expression(paste("Literacy skills at  ", T[1])),
    expression(paste("Numeracy skills at  ", T[1]))
  )

  plotlims <- c(
    ifelse(grepl("piaac", data),
      500, 4
    ),
    ifelse(grepl("piaac", data),
      0,
      -4
    )
  )

  plotdata <- get(data) %>%
    filter(.imp < 6) %>%
    mutate(PV = as_factor(.imp)) 

  pointplot <- plotdata %>%
    ggplot(aes(
      y = t2_pv, x = t1_pv,
      group = PV, color = PV, fill = PV
    )) +
    geom_point(size = 0.1) +
    geom_smooth(method = "lm", se = FALSE,
                size = 0.2) +
    scale_color_brewer("PV", palette = "Blues") +
    scale_fill_brewer("PV", palette = "Blues") +
    theme_apa() +
    expand_limits(y = plotlims, x = plotlims) +
    ylab(y_lab) +
    xlab(x_lab) +
    theme(
      text = element_text(family = "Source Sans Pro",
                          size = 10),
      axis.title = element_text(size = 10),
      axis.text = element_text(size = 10),
      plot.margin = margin(1, 0.5, 0.5, 0.5, "cm")
    )

  pointplot
}

points_plot <- cowplot::plot_grid(
  point_plotter("reading_piaac"),
  point_plotter("reading_neps"),
  point_plotter("math_piaac"),
  point_plotter("math_neps"),
  nrow = 2,
  labels = c("Three years (PIAAC-L)", "Six years (NEPS)"),
  align = "h",
  hjust = -0.1,
  label_fontfamily = "Source Sans Pro",
  label_size = 12
)

points_plot
```
