# Supplementary Tables and Figures

(ref:tableA1-caption) Skills at $T_1$ and skill change from $T_1$ to $T_2$ in the two samples when survey weights are applied.

```{r tableA1, echo = FALSE, warning = FALSE}

# Options for survey package


# Merge information about change in literacy and numeracy per study
# i.e., combine the information for both domains
# Compute pooled correlations between change (delta) and initial level (t1)

# Combine descriptive statistics about change in a reduced and formatted table
skills_descriptives <- deltas %>%
  filter(grouping == "total") %>%
  select(target, t1, sd_t1, t2, sd_t2, delta, sd_delta) %>%
  arrange(-grepl("piaac", target)) %>%
  pivot_longer(!target) %>%
  mutate(
    study = str_extract(target, "piaac|neps"),
    domain = str_extract(target, "math|reading")
  ) %>%
  select(-target) %>%
  pivot_wider(names_from = c(study, domain), values_from = c(value)) %>%
  mutate(name = str_replace_all(
    name,
    c(
      "sd_t[12]|sd_delta" = "\\\\emph{SD}",
      "t[12]_{0}|delta" = "\\\\emph{M}"
    )
  )) %>%
  select(
    Parameter = name,
    "Literacy" = piaac_reading,
    "Numeracy" = piaac_math,
    "Reading" = neps_reading,
    "Math" = neps_math
  )

papaja::apa_table(skills_descriptives,
  col_spanners =
    list("\\makecell[c]{PIAAC-L\\\\(3 years)}" = c(2, 3), "\\makecell[c]{NEPS\\\\(6 years)}" = c(4, 5)),
  stub_indents = list(
    "Skill level at $T_1$" = c(1:2),
    "Skill level at $T_2$" = c(3:4),
    "Skill change ($\\Delta{T_1, T_2}$)" = c(5:6)
  ),
  escape = F,
  caption = "(ref:tableA1-caption)",
  font_size = "small",
  note = "Statistics are based on weighted data (see Table 1 in the manuscript for unweighted statistics). Means are in the raw metric of the tests (0-500 points in PIAAC; logits in NEPS). All statistics are pooled across 10 (30) PVs in PIAAC-L (NEPS)."
)

```

\newpage

(ref:figureA1-caption) Scatterplots of literacy and numeracy skills at the two time points. For better legibility, the plot shows data from a single PV. Axes are in the original metrics of the PIAAC (0-500 points) and NEPS tests (logits), respectively. The color of the points corresponds to the survey weight of the observation (darker shades represent higher weights). The lines show the linear correlation coefficient $r_{T_1, T_2}$. 

```{r figureA1, echo = FALSE, warning = FALSE, fig.dim = c(9,7), fig.cap = "(ref:figureA1-caption)"}
point_plotter <- function(data, use_pv = 5) {
  y_lab <- if_else(grepl("reading", data),
                   expression(paste("Literacy skills at  ", T[2])),
                   expression(paste("Numeracy skills at  ", T[2]))
  )
  x_lab <- if_else(grepl("reading", data),
                   expression(paste("Literacy skills at  ", T[1])),
                   expression(paste("Numeracy skills at  ", T[1]))
  )
  
  plotlims <- c(
    ifelse(grepl("piaac", data),
           450, 3
    ),
    ifelse(grepl("piaac", data),
           50,
           -3
    )
  )
  
  plotdata <- get(data) %>%
    filter(.imp == use_pv) %>% # Select only a single PV
    mutate(PV = as_factor(.imp),
           plotweight = ntile(weight, 9) %>% as_factor()
      
           ) 

  pointplot <- plotdata %>%
    ggplot(aes(
      y = t2_pv, x = t1_pv,
      color = plotweight, fill = plotweight 
      #group = PV, 
    )) +
    geom_point(size = 0.3) +
    geom_smooth(aes(y = t2_pv, x = t1_pv), method = "glm", se = FALSE, 
                color = "#CC0001", inherit.aes = FALSE,
                size = 0.3) +
    scale_color_brewer(palette = "Blues") +
    scale_fill_brewer(palette = "Blues") +
    theme_apa() +
    ylab(y_lab) +
    xlab(x_lab) +
    theme(
      text = element_text(#family = "Source Sans Pro",
        size = 10),
      axis.title = element_text(size = 10),
      axis.text = element_text(size = 10),
      plot.margin = margin(1, 0.5, 0.5, 0.5, "cm"),
      legend.position = "none"
    )  + 
    expand_limits(y = plotlims, x = plotlims)  
  
  pointplot
  
}

 plot1 <- point_plotter("reading_piaac")
 plot2 <-  point_plotter("reading_neps")
 plot3 <- point_plotter("math_piaac")
 plot4 <- point_plotter("math_neps")


points_plot <- cowplot::plot_grid(
  plot1, plot2, plot3, plot4,
  nrow = 2,
  labels = c("Three years (PIAAC-L)", "Six years (NEPS)"),
  align = "h",
  hjust = -0.1,
  # label_fontfamily = "Source Sans Pro",
  label_size = 12
)

points_plot
```

\newpage

(ref:figureA2-caption) Cross-sectional age profiles of literacy and numeracy skills at $T_1$. The horizontal line represents the $T_1$ sample mean. The *Y* axis shows the raw metric of the tests (0-500 points in PIAAC; logits in NEPS). The axis ranges are chosen to include 1.5\ *SD* below and above the sample mean. The figure is based on a single PV per skill and study. It uses survey weights and a bandwidth of 5 for the smoothing function.

```{r figureA2, echo = FALSE, warning = FALSE, fig.dim = c(8,7), fig.cap = "(ref:figureA2-caption)"}


# Plot options
use_bandwidth <- 5

# Plot function
ageplot <- function(input_data, weight = use_weight, bandwidth = use_bandwidth ) {
  
  # Get the data
  data <- base::get(input_data) %>% 
    filter(.imp == 5)
  
  
  design <- data %>%
    svydesign(data = ., 
              ids = ~psu, strata = ~stratum, weights = weight, #use cross-sectional weight
               nest = TRUE)
  
  
  plotdata <- svysmooth(t1_pv ~ age, bandwidth = bandwidth,
            design = design) 
  

   t1stats <- deltas %>%
    filter(target == input_data & grouping == "total") %>%
    select(t1, sd_t1)
  
   
   domain <- if_else(grepl("reading", input_data), "Literacy", "Numeracy")
   
   
   plotlims <- list(
     upper = t1stats$t1 + 1.5*t1stats$sd_t1
     ,
     lower = t1stats$t1 - 1.5*t1stats$sd_t1
    )
   
   plotresult <- plotdata %>% 
     pluck("age") %>%
     as_tibble() %>%
     ggplot(aes(x = x, y = y)) + 
     geom_hline(aes(yintercept = t1stats$t1), color = "#666666") +
     geom_line(color = "#005B96", size = 1) +
     expand_limits(
       y = c(plotlims$lower, plotlims$upper),
       x = c(18, 67)
     ) +
   
     xlab(expression(paste("Age at ", T[1], " in years"))) +
     ylab(expression(paste("Skill level at ", T[1]))) +
     # scale_y_continuous(breaks =
     #                     ifelse(grepl("piaac", data),
     #                     seq(250, 300),
     #                     seq(-1, 1))
     #                     ) +
     theme(
       # text = element_text(family = "Source Sans Pro"),
       panel.background = element_rect(fill = "white", color = "black"),
       axis.title.x = element_text(size = 10, margin = margin(t = 10)),
       axis.title.y = element_text(size = 10, margin = margin(r = 10)),
       axis.text = element_text(size = 10),
       axis.line = element_line(color = "black"),
    #   panel.grid.major = element_line(colour = "grey90", size = 0.1),
       plot.margin = margin(1, 0.5, 0.5, 0.5, "cm"),
       panel.grid = element_blank()
     )
   
   
  plotresult
  
}


# Plot the result in a 2x2 panel
cowplot::plot_grid(
  ageplot("reading_piaac"),
  ageplot("reading_neps"),
  ageplot("math_piaac"),
  ageplot("math_neps"),
  labels = c(
    "Literacy (PIAAC-L)", "Literacy (NEPS)",
    "Numeracy (PIAAC-L)", "Numeracy (NEPS)"
  ),
  label_size = 12,
  #  label_fontfamily = "Source Sans Pro",
  label_y = 1
)

```


\newpage

(ref:tableA2-caption) Linear age differences in literacy and numeracy skills as estimated from the cross-sectional age profiles at $T_1$ (based on weighted data).

```{r tableA2, echo = F, warning = F}
# Table is sourced from 04_reporting.R
# source("04_reporting.R")

age_effects %>%
  mutate(Change = str_replace_all(
    Change,
    "SD",
    "\\\\emph{SD}"
  )) %>%
  rename("Linear age effect" = Change) %>%
  apa_table(
    col.names = c("Linear age effect", rep(c("Literacy", "Numeracy"), 2)),
    col_spanners =
      list(
        "PIAAC-L" = c(2, 3),
        "NEPS" = c(4, 5)
      ),
    stub_indents = list(
      "For one year" = c(1:2),
      "For the study period" = c(3:4),
      "For one decade" = c(5:6)
    ),
    escape = F,
    caption = "(ref:tableA2-caption)",
    align = "lcccc",
    font_size = "small",
    note = "Cross-sectional age differences in skills  estimated from a linear model regressing $T_1$ skill level on age (based on weighted data). The study period was three years in PIAAC-L and six years in NEPS. \"$SD$\" refers to change expressed in pooled standard deviations of the respective skill. All estimates are pooled across the 10 (30 in NEPS) PVs."
  )
```


\newpage

 
(ref:tableA3-caption) Mean-level change depending on the initial skill level ($T_1$ skill quartile).

```{r tableA3, echo = F, warning = F}

deltas.table %>% 
  filter(str_detect(Group, "quarti")) %>% 
  select(Group, starts_with("dav")) %>% 
   apa_table(
    col.names = c("Skill level at $T_1$", rep(c("Literacy", "Numeracy"), 2)),
    col_spanners =
      list(
        "PIAAC-L" = c(2, 3),
        "NEPS" = c(4, 5)
      ),
    escape = F,
    caption = "(ref:tableA2-caption)",
    align = "lcccc",
    font_size = "small",
    note = "Effect sizes ($d_{av}$) for mean-level change in skills depending on initial skill level (i.e., $T_1$ skill quartile, based on weighted data). All estimates are pooled across the 10 (30 in NEPS) PVs."
  )
  

```
\newpage

