
# Supplementary Tables and Figures.

```{r setup2, include = FALSE}
# Load required packages

library(kableExtra)
library(extrafont) 
library(papaja)
library(tidyverse)
library(glue)
library(sjlabelled)
library(mice)
library(miceadds)
library(mitools)
library(srvyr)
library(cowplot)

# BibLateX Reference file
r_refs("r-references.bib")

# Load raw data and results
map(c(
  "math_neps.Rda", "reading_neps.Rda",
  "math_piaac.Rda", "reading_piaac.Rda",
  "./02_results/cors.Rda",
  "./02_results/deltas.Rda"
), load, .GlobalEnv)
```

\newpage

(ref:tableS1-caption) Skills and skill change in the two sample (after weighting).

```{r tableS1, echo = FALSE}
# Merge information about change in literacy and numeracy per study
# i.e., combine the information for both domains
# Compute pooled correlations between change (delta) and initial level (t1)
cors_pooler2 <- function(target) {
  data_filtered <- get(target) %>%
    split(.$.imp) %>%
    imputationList()

  sink("NUL")

  results <- with(data_filtered, lm(scale(t2_pv - t1_pv) ~ scale(t1_pv),
    weights = weight
  )) %>%
    MIcombine() %>%
    summary()

  sink()

  resultlist <- tibble(
    rho = results[["results"]][2],
    se = results[["se"]][2],
    lower = results[["(lower"]][2],
    upper = results[["upper)"]][2]
  )

  resultlist
}

# Create a tibble for computing the PV-based correlations of t1 with delta
cors_t1_delta <- tibble(
  target = c("reading_neps", "math_neps", "reading_piaac", "math_piaac"),
)

# Map the function to the tibble to obtain the results
cors_t1_delta <- cors_t1_delta %>%
  rowwise() %>%
  mutate(ergebnis = list(cors_pooler2(target))) %>%
  unnest(ergebnis) %>%
  mutate(
    domain = str_extract(target, pattern = "reading|math"),
    study = str_extract(target, pattern = "neps|piaac"), 
    CI = str_c("[", round(lower,2),"; ", round(upper, 2), "]")
  )

# Combine descriptive statistics about change in a reduced and formatted table
skills_descriptives <- deltas %>%
  filter(grouping == "total") %>%
  select(target, t1, sd_t1, t2, sd_t2, delta, sd_delta) %>%
  arrange(-grepl("piaac", target)) %>%
  pivot_longer(!target) %>%
  mutate(
    study = str_extract(target, "piaac|neps"),
    domain = str_extract(target, "math|reading")
  ) %>%
  select(-target) %>%
  pivot_wider(names_from = c(study, domain), values_from = c(value)) %>%
  mutate(name = str_replace_all(
    name,
    c(
      "sd_t[12]|sd_delta" = "\\\\emph{SD}",
      "t[12]_{0}|delta" = "\\\\emph{M}"
    )
  )) %>%
  select(
    Parameter = name,
    "Literacy" = piaac_reading,
    "Numeracy" = piaac_math,
    "Reading" = neps_reading,
    "Math" = neps_math
  )

apa_table(skills_descriptives,
  col_spanners =
    list("\\makecell[c]{PIAAC-L\\\\(3 years)}" = c(2, 3), "\\makecell[c]{NEPS\\\\(6 years)}" = c(4, 5)),
  stub_indents = list(
    "Skill level at $T_1$" = c(1:2),
    "Skill level at $T_2$" = c(3:4),
    "Skill change ($\\Delta{T_1, T_2}$)" = c(5:6)
  ),
  escape = F,
  caption = "(ref:tableS1-caption)",
  font_size = "small",
  note = "All values are pooled across 10 (20) PVs in PIAAC-L (NEPS)."
)
```

\newpage

(ref:figureS1-caption) Cross-sectional age profiles of literacy and numeracy skills at $T_1$. The gray horizontal line represents the sample mean. 

```{r figureS1, echo = FALSE, warning = FALSE, fig.cap = "(ref:figureS1-caption)"}

ageplot <- function(data) {

  # Compute skill mean, min and max at T1 - then pool the data across PVs

  plotdata <- get(data)
  t1skills <- plotdata %>%
    group_by(.imp) %>%
    summarise(
      Mean = mean(t1_pv),
      Min = min(t1_pv),
      Max = max(t1_pv)
    ) %>%
    summarise(across(where(is.numeric), mean))


  domain <- if_else(grepl("reading", data), "Literacy", "Numeracy")

  plotlims <- list(
  upper = ifelse(grepl("piaac", data),
        300, 1),
  lower = ifelse(grepl("piaac", data),
        250,
        -1)
    )
  
  # Plot the age profile based on the first PV
  ageplot <- plotdata %>%
    filter(.imp == 1) %>%
    ggplot(aes(x = age, y = t1_pv)) +
    expand_limits(
       y = c(plotlims$lower, plotlims$upper),
      x = c(18, 67)
    ) +
    geom_smooth(
      #span = 0.1, level = 0.95, 
      fill = "#58748F", color = "#58748F"
    ) +

    # stat_smooth(level = 0.99, span = 0.01) +
    geom_hline(aes(yintercept = t1skills[["Mean"]]), color = "#666666") +
    xlab(expression(paste("Age at ", T[1], " in years"))) +
    ylab(expression(paste("Skill level at ", T[1]))) +
    # scale_y_continuous(breaks =
    #                     ifelse(grepl("piaac", data),
    #                     seq(250, 300),
    #                     seq(-1, 1))
    #                     ) +
    theme(
      text = element_text(family = "Source Sans Pro"),
      panel.background = element_rect(fill = "white", color = "black"),
      axis.title.x = element_text(size = 12, margin = margin(t = 10)),
      axis.title.y = element_text(size = 12, margin = margin(r = 10)),
      axis.text = element_text(size = 12),
      axis.line = element_line(color = "black"),
      panel.grid.major = element_line(colour = "grey90", size = 0.1),
      plot.margin = margin(1, 0.5, 0.5, 0.5, "cm")
    )

  ageplot
}

plot_grid(
  ageplot("reading_piaac"),
  ageplot("reading_neps"),
  ageplot("math_piaac"),
  ageplot("math_neps"),
  labels = c(
    "Literacy (PIAAC-L)", "Literacy (NEPS)", 
     "Numeracy (PIAAC-L)", "Numeracy (NEPS)"
  ),
  label_size = 15,
  label_fontfamily = "Source Sans Pro",
  label_y = 1
)
```

\newpage
(ref:figureS2-caption) Distribution of the change scores ($\Delta{T_1, T_2)}$. The gray horizontal line represents "no change". The dark blue areas represent change greater than 0.8 *SD*, which is conventionally seen as a "large" effect. 

```{r figureS2, echo = FALSE, warning = FALSE, fig.cap = "(ref:figureS2-caption)"}

delta_density <- function(data) {
  
  data <- get(data)
  # Compute sd at T1 (for computation of Cohen's dav and for coloring the plot)
  require(radiant.data)
  
  sds <- data %>%
    group_by(.imp) %>%
    summarise(
      sd_pooled = 0.5 * (weighted.sd(t1_pv, weight) +
                           weighted.sd(t2_pv, weight)),
      sd_delta = weighted.sd(t2_pv - t1_pv, weight)
    ) %>%
    summarise(sd_pooled = mean(sd_pooled), sd_delta = mean(sd_delta))
  
  # Compute deltas (t2-t1) based on the EAP scores
  plotdata <- data %>% ungroup %>%
    # group_by(seqid) %>%
    mutate(delta = t1_pv - t2_pv)
    
    # Compute densities for each delta value and categorize them according to  whether they exeed a "large" effect.
  plotdata <- with(density(plotdata$delta), tibble(x, y)) %>%
    mutate(
      direction = ifelse(x <= -0.8 * sds[["sd_pooled"]], "neg",
                         ifelse(x >= 0.8 * sds[["sd_pooled"]], "pos", "little")
      )
    )
  
  # Create a density plot
 plotted <- plotdata %>% ggplot(aes(x = x, y = y)) +
    geom_area(aes(fill = direction),
              alpha = 1,
              show.legend = F
    ) +
    geom_vline(
      xintercept = 0,
      size = 0.5,
      lty = "dashed",
      color = "#666666"
    ) +
    geom_line() +
    theme_apa() +
    xlab(expression(paste("Change score (", Delta, "T2, T1", ")"))) +
    ylab("Density") +
    theme(
      text = element_text(family = "Source Sans Pro"),
      axis.title = element_text(size = 12),
      axis.text = element_text(size = 12),
      plot.margin = margin(1, 0.5, 0.5, 0.5, "cm")
    ) +
    scale_fill_manual(values = c("#BDCDD8", "#58748F", "#58748F")) +
    expand_limits(x = c(-2 * sds$sd_pooled, 2 * sds$sd_pooled))
  
 return(plotted)
 
}

plot_grid(
  delta_density("reading_piaac"),
  delta_density("reading_neps"),
  delta_density("math_piaac"),
  delta_density("math_neps"),
   labels = c(
    "Literacy (PIAAC-L)", "Literacy (NEPS)", 
     "Numeracy (PIAAC-L)", "Numeracy (NEPS)"
  ),
  label_size = 15,
  label_fontfamily = "Source Sans Pro",
  label_y = 1
  )

```